---
title: "AAQ-2 Psychometric Analysis with Rasch Measurement Theory"
title-block-banner: "#009ca6"
title-block-banner-color: "#FFFFFF"
author: 
  name: Magnus Johansson
  affiliation: RISE Research Institutes of Sweden
  affiliation-url: https://www.ri.se/en/kbm
  orcid: 0000-0003-1669-592X
date: last-modified
date-format: iso
repo-url: https://github.com/pgmj/aaq2_rasch
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of contents"
    embed-resources: true
    standalone: true
    page-layout: full
    mainfont: 'Lato'
    monofont: 'Roboto Mono'
    code-overflow: wrap
    code-fold: true
    code-tools: true
    number-sections: true
    fig-dpi: 96
    layout-align: left
    linestretch: 1.6
    theme:
      - materia
      - custom.scss
    css: styles.css
    license: CC BY
execute:
  echo: true
  warning: false
  message: false
  cache: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
bibliography: 
  - grateful-refs.bib
  - references.bib
---

```{r}
#| label: setup

# one package below requires that you use devtools to install it manually:
# first install devtools by
# install.packages('devtools')

library(RISEkbmRasch) # devtools::install_github("pgmj/RISEkbmRasch")
library(grateful)
library(ggrepel)
library(car)
library(kableExtra)
library(readxl)
library(tidyverse)
library(eRm)
library(mirt)
library(psych)
library(psychotree)
library(matrixStats)
library(reshape)
library(knitr)
library(patchwork)
library(formattable) 
library(glue)
library(haven)
library(labelled)

### optional libraries
#library(TAM)
#library(skimr)

### some commands exist in multiple packages, here we define preferred ones that are frequently used
select <- dplyr::select
count <- dplyr::count
recode <- car::recode
rename <- dplyr::rename
```

## Background

Data from [@langer2024]. More comments on that paper will be added
later. The only other Item Response Theory-based paper found analyzing
the AAQ-2 is [@ong2019] but they do a rather limited analysis using the
Graded Response Model, for example omitting test of unidimensionality,
local independence of items, and ordering of response categories.

```{r}
### import data
df.all <- read_spss("data/DataBase AAQ-II_Criteria_rev.sav")

df <- df.all %>% 
  select(starts_with("AAQ_II"),GenderBirth)

dif.sex <- df$GenderBirth
df$GenderBirth <- NULL
names(df) <- paste0("aaq_",c(1:7))

#val_labels(dif.sex)

itemlabels <- data.frame(
  stringsAsFactors = FALSE,
  itemnr = paste0("aaq_",c(1:7)),
  item = c("My painful experiences and memories make it difficult for me to live a life that I would value",
    "I’m afraid of my feelings",
    "I worry about not being able to control my worries and feelings",
    "My painful memories prevent me from having a fulfilling life",
    "Emotions cause problems in my life",
    "It seems like most people are handling their lives better than I am",
    "Worries get in the way of my success"
  )
)

source("RISE_theme.R")
```

## All items in the analysis

```{r}
RIlistitems(df)
```

## Demographics

```{r}
#| layout-ncol: 2
RIdemographics(dif.sex, "Gender")
```

```{r}
hist(df.all$Age, title = "Age distribution", xlab = "Age", main = "Age distribution", col = "lightblue")
#abline(v = mean(df.all$Age), col = "red", lwd = 2)
```

Very young sample, and not a lot of variation in age either.

## Descriptives of raw data

Response distribution for all items together.

```{r}
#| tbl-cap: "Total number of responses for all items"
RIallresp(df)
```

We need the response scale to start at 0 instead of 1 for the Rasch
analysis to work correctly.

```{r}
df <- df %>% 
  mutate(across(starts_with("aaq_"), ~ .x - 1))
```

```{r}
allresp <- data.frame(
    Response.category = c(0L, 1L, 2L, 3L, 4L, 5L, 6L, NA),
  Number.of.responses = c(2541L, 2263L, 1443L, 2241L, 1525L, 1046L, 857L, 19L),
              Percent = c(21.3, 19, 12.1, 18.8, 12.8, 8.8, 7.2, 0.2)
)

ggplot(allresp, aes(x = Response.category, y = Percent)) +
  geom_bar(stat = "identity", fill = "#009ca6") +
  geom_text(aes(label = Number.of.responses), vjust = -0.5, size = 3) +
  labs(title = "Response distribution for all items",
       x = "Response category",
       y = "% of responses") +
  scale_x_continuous(breaks = c(0:6)) +
  theme_rise()

```


### Descriptives - item level

```{r}
#| column: margin
#| echo: false
RIlistItemsMargin(df, fontsize = 12)
```

::: panel-tabset
#### Tile plot

```{r}
RItileplot(df)
```

#### Stacked bars

```{r}
RIbarstack(df)
```

#### Barplots

```{r}
#| layout-ncol: 2
RIbarplot(df)
```
:::

Data is skewed towards the lower end of the scale. Most item responses
are a bit oddly distributed. Response category 2 is consistently
deviating from the expected (normal-ish) pattern. Makes one wonder about
the data collection and the response category wording used. Let us
check.

```{r}
val_labels(df.all$AAQ_II_1)
```

My Spanish is not great, so here is the Google Translated version:

1.  It's never true for me

2.  It's very rarely true for me.

3.  It's rarely true for me

4.  Sometimes it's true for me

5.  It's often true for me.

6.  It's almost always true for me.

7.  It's always true for me

### Missing data

::: panel-tabset
#### Missing responses/item

```{r}
RImissing(df)
```

#### Missing responses/person

```{r}
RImissingP(df, n = 20)
```
:::

One respondent with 100% missing, and 12 with 1 item missing. We'll
remove all respondents with missing data since we have a large dataset.

```{r}
df <- na.omit(df)
dif.sex <- df.all %>% 
  select(GenderBirth,starts_with("AAQ")) %>%
  na.omit() %>% 
  pull(GenderBirth)

dif.sex <- factor(dif.sex, levels = c(1,2),
                  labels = c("Male","Female"))

```

## Rasch analysis 1

The eRm package, which uses Conditional Maximum Likelihood (CML)
estimation, will be used primarily. For this analysis, the Partial
Credit Model will be used.

```{r}
#| column: margin
#| echo: false
RIlistItemsMargin(df, fontsize = 13)
```

::: panel-tabset
### Item fit

```{r}
RIitemfitPCM2(df, 250, 16
              )
```

### PCA

```{r}
#| tbl-cap: "PCA of Rasch model residuals"
RIpcmPCA(df)
```

### Residual correlations

```{r}
RIresidcorr(df, cutoff = 0.2)
```

### 1st contrast loadings

```{r}
RIloadLoc(df)
```

### Response categories

```{r}
#| layout-ncol: 2
RIitemCats(df)
```

### Targeting

```{r}
#| fig-height: 5
# increase fig-height above as needed, if you have many items
RItargeting(df)
```

### Item hierarchy

```{r}
#| fig-height: 5
RIitemHierarchy(df)
```

### DIF 1

```{r}
erm.df <- PCM(df)
LRtest(erm.df, dif.sex)
RIdifTableLR(df, dif.sex)
```
:::

Item 5 has somewhat low item fit. PCA of residuals is below 2, but
residual correlations show several issues. The strongest correlation is
between items 1 and 4, followed by 2 and 3. Items 5 and 7 are also above
the cutoff.

The tests for DIF indicates possible issues (p \< .05), but the DIF
table shows that the differences are small, at least on the item average level. The p-value is small due to
the large sample.

All items show disordered response categories. The pattern is most
clearly seen in the targeting figure. The consistency observed, with
threshold 3 below threhold 2 for all items, makes one wonder if there
was a mistake in the coding of responses to numbers, but we don't know
if the supplied datafile is the original.

We'll merge categories 1 and 2:

-   It's very rarely true for me.
-   It's rarely true for me

```{r}
df2 <- df %>% 
  mutate(across(starts_with("aaq_"), ~ recode(., "2=1;3=2;4=3;5=4;6=5")))

```

### 1 and 2 merged

```{r}
mirt(df2, model=1, itemtype='Rasch', verbose = FALSE) %>% 
  plot(type="trace", as.table = TRUE, 
       theta_lim = c(-6,6))
```

Let's also try merging 2 and 3 for comparison.

```{r}
df2 <- df %>% 
  mutate(across(starts_with("aaq_"), ~ recode(., "3=2;4=3;5=4;6=5")))

```

### 2 and 3 merged instead

```{r}
mirt(df2, model=1, itemtype='Rasch', verbose = FALSE) %>% 
  plot(type="trace", as.table = TRUE, 
       theta_lim = c(-6,6))
```

It seems 1 and 2 works better, as expected.

```{r}
df2 <- df %>% 
  mutate(across(starts_with("aaq_"), ~ recode(., "2=1;3=2;4=3;5=4;6=5")))
```

### Residual correlations

We have several item pairs that correlate above the relative cutoff of
0.2 and will deal with them one at a time. The strongest correlation is
between items 1 and 4.

-   item 1: My painful experiences and memories make it difficult for me
    to live a life that I would value
-   item 4: My painful memories prevent me from having a fulfilling life

These items are very much alike, apart from "fulfilling life" vs "a life
that I would value" and item 1 adding "experiences" to "memories", but
even these differences are very similar. It's not surprising that they
correlate strongly.

Item 1 has better separation of item response thresholds. We'll remove
item 4.

```{r}
df2 <- df2 %>% 
  select(!aaq_4)
```

## Rasch analysis 2

```{r}
#| column: margin
#| echo: false
RIlistItemsMargin(df, fontsize = 13)
```

::: panel-tabset
### Item fit

```{r}
RIitemfitPCM2(df2, 250, 16)
```

### PCA

```{r}
#| tbl-cap: "PCA of Rasch model residuals"
RIpcmPCA(df2)
```

### Residual correlations

```{r}
RIresidcorr(df2, cutoff = 0.2)
```

### 1st contrast loadings

```{r}
RIloadLoc(df2)
```

### Targeting

```{r}
#| fig-height: 5
# increase fig-height above as needed, if you have many items
RItargeting(df2)
```

### Item hierarchy

```{r}
#| fig-height: 5
RIitemHierarchy(df2)
```
:::

Items 2 and 3 still correlate quite strongly.

-   item 2: I'm afraid of my feelings
-   item 3: I worry about not being able to control my worries and
    feelings

Item 3 has worse fit, and we'll remove it. It is also a dual question,
"worries and feelings", which can explain the item fit.

```{r}
df2 <- df2 %>% 
  select(!aaq_3)
```

## Rasch analysis 3

```{r}
#| column: margin
#| echo: false
RIlistItemsMargin(df, fontsize = 13)
```

::: panel-tabset
### Item fit

```{r}
RIitemfitPCM2(df2, 250, 16)
```

### PCA

```{r}
#| tbl-cap: "PCA of Rasch model residuals"
RIpcmPCA(df2)
```

### Residual correlations

```{r}
RIresidcorr(df2, cutoff = 0.2)
```

### 1st contrast loadings

```{r}
RIloadLoc(df2)
```

### Targeting

```{r}
#| fig-height: 5
# increase fig-height above as needed, if you have many items
RItargeting(df2)
```

### Item hierarchy

```{r}
#| fig-height: 5
RIitemHierarchy(df2)
```
:::

There is a large gap in targeting due to the dysfunctional response
categories that needed merging. Item 5 is low in item fit, "Emotions
cause problems in my life". It is a very general item, which may explain
the low fit. We'll keep it for now.

## DIF-analysis

### Gender

```{r}
#| column: margin
#| echo: false
RIlistItemsMargin(df2, fontsize = 13)
```

::: panel-tabset
#### Average locations

```{r}
RIdifTableLR(df2, dif.sex)
RIdifFigureLR(df2, dif.sex)
```

#### Threshold locations

```{r}
RIdifThreshTblLR(df2, dif.sex)
RIdifThreshFigLR(df2, dif.sex)
```
:::

Item 1 shows a surprising pattern for the two highest threholds.

## Other metrics

### Reliability

```{r}
RItif(df2)
```

### Person location & infit ZSTD

```{r}
RIpfit(df)
```

### Item parameters

```{r}
RIitemparams(df2)
```

### Transformation table

```{r}
RIscoreSE(df2)
```

### Ordinal/interval figure

```{r}
RIscoreSE(df2, output = "figure")
```

## Summary of Rasch analysis

Two item pairs had strongly correlated residuals, and all items had
issues with disordered thresholds related to the second lowest response
category. There was no DIF for gender at birth.

Targeting is not great. Items have rather similar locations, and there
is a large gap where the average person location is. If the intended use
is clinical, there is decent reliability for those with above average
locations.

## Comparison of latent scores

The "standard" way to use the AAQ-2 is to sum/average the items, after
recoding item categories to numerics. We will compare this to the Rasch
estimated latent scores.

For ordinal sum scores, we will use the original 7 items with their
original 7 response categories, even though the analysis does not
support this. We do this to illustrate the difference that can be hidden
behind data that is unjustifiedly sum scored and data based on
psychometric analysis.

```{r}
df.viz <- df %>% 
  na.omit() %>% 
  mutate(sumscore = rowSums(across(starts_with("aaq_")))) %>% 
  mutate(sumscore = as.numeric(sumscore)) %>% 
  rownames_to_column("id") %>% 
  mutate(id = as.numeric(id))

df.viz$latentscore <- RIestThetas2(df2)
```

```{r}
ggplot(df.viz, aes(x = latentscore, y = sumscore)) +
  geom_point() +
  labs(x = "Rasch estimated latent score", y = "Ordinal sum score") +
  theme_rise()
```

```{r}
pfit <- RIpfit(df2, output = "dataframe") %>% 
  janitor::clean_names() %>% 
  rename(id = rownumber)

df.viz <- left_join(df.viz, pfit, by = "id")

df.viz <- df.viz %>% 
  mutate(person_fit = ifelse(person_infit_zstd > 2 | person_infit_zstd < -2, "outlier", "normal"))

ggplot(df.viz, aes(x = latentscore, y = sumscore, color = person_fit)) +
  geom_point() +
  labs(x = "Rasch estimated latent score", y = "Ordinal sum score") +
  theme_rise() +
  scale_color_viridis_d() +
  geom_smooth(method = "loess", color = "darkgrey")

```

```{r}
hist(df.viz$latentscore, breaks = 20, main = "Histogram of Rasch estimated (WL) latent scores", col = "lightpink")
```

```{r}
hist(df.viz$sumscore, breaks = 20, main = "Histogram of ordinal sum scores", col = "lightblue")
mtext(text = "Note: this is based on the original 7 items and 7 response categories.", 
      side = 3, line = 0.4)
```

## CFA

Let's make a brief comparison.

```{r}
library(lavaan)
library(lavaanExtra)
```

Here is some info about the CFA methodology:
<https://pgmj.github.io/ki_irt_mhcsf/cfa.html> It will be copied to this
document later.

We will use both WLSMV and ML/MLR estimators.

```{r}
# Define latent variables
latent <- list(
  aaq2 = names(df)
)

# Write the model
cfa.model <- write_lavaan(latent = latent)

fit.wlsmv <- cfa(model = cfa.model, 
               data = df,
               estimator = "WLSMV",
               ordered = TRUE,
               rotation = "oblimin")

fit.ml <- cfa(model = cfa.model, 
               data = df,
               estimator = "ML",
               rotation = "oblimin")

fit.mlr <- cfa(model = cfa.model, 
               data = df,
               estimator = "MLR",
               rotation = "oblimin")

```

```{r}
fit_metrics_robust <- c("chisq.scaled", "df", "pvalue.scaled", 
                         "cfi.robust", "tli.robust", "rmsea.robust", 
                        "rmsea.ci.lower.robust","rmsea.ci.upper.robust",
                        "srmr")

mfit1 <- 
  fitmeasures(fit.wlsmv, fit_metrics_robust) %>% 
  rbind() %>% 
  as.data.frame() %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>%
  rename(Chi2 = chisq.scaled,
         p = pvalue.scaled,
         CFI = cfi.robust,
         TLI = tli.robust,
         RMSEA = rmsea.robust,
         CI_low = rmsea.ci.lower.robust,
         CI_high = rmsea.ci.upper.robust,
         SRMR = srmr) %>% 
  add_column(Model = "WLSMV", .before = "Chi2")
```

```{r}
fit_metrics.ml <- c("chisq", "df", "pvalue", 
                  "cfi", "tli", "rmsea", 
                  "rmsea.ci.lower","rmsea.ci.upper",
                  "srmr")

mfit2 <- 
  fitmeasures(fit.ml, fit_metrics.ml) %>% 
  rbind() %>% 
  as.data.frame() %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>% 
    rename(Chi2 = chisq,
         p = pvalue,
         CFI = cfi,
         TLI = tli,
         RMSEA = rmsea,
         CI_low = rmsea.ci.lower,
         CI_high = rmsea.ci.upper,
         SRMR = srmr) %>% 
  add_column(Model = "ML", .before = "Chi2")

```

```{r}
fit_metrics.mlr <- c("chisq.scaled", "df.scaled", "pvalue", 
                  "cfi.scaled", "tli.scaled", "rmsea.scaled", 
                  "rmsea.ci.lower.scaled","rmsea.ci.upper.scaled",
                  "srmr")

mfit3 <- 
  fitmeasures(fit.mlr, fit_metrics.mlr) %>% 
  rbind() %>% 
  as.data.frame() %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>% 
    rename(Chi2 = chisq.scaled,
         p = pvalue,
         df = df.scaled,
         CFI = cfi.scaled,
         TLI = tli.scaled,
         RMSEA = rmsea.scaled,
         CI_low = rmsea.ci.lower.scaled,
         CI_high = rmsea.ci.upper.scaled,
         SRMR = srmr) %>% 
  add_column(Model = "MLR", .before = "Chi2")

```

### Model fit comparison

```{r}
mfit1 %>% 
  bind_rows(mfit2) %>% 
  bind_rows(mfit3) %>% 
  rownames_to_column("remove") %>%
  select(!remove) %>%
  kbl_rise()
```

### Modification indices

We'll look at the MLR output, since that model had the least bad fit.

```{r}
modificationIndices(fit.mlr,
                    standardized = T) %>% 
  as.data.frame(row.names = NULL) %>% 
  filter(mi > 30,
         op == "~~") %>% 
  arrange(desc(mi)) %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>%
  kbl_rise(fontsize = 14, tbl_width = 75)
```

Huge residual correlations between items 1 and 4, and 2 and 3.
Amusingly, GitHub Copilot now suggests to me that we add these to the
model, which is a really bad practice that too often is used.
Unidimensionality is a key assumption of this CFA, and we should not add
items to the model just because they have high residual correlations.

### CFA based on Rasch model

```{r}
# Define latent variables
latent <- list(
  aaq2 = names(df2)
)

# Write the model
cfa.model <- write_lavaan(latent = latent)

fit.wlsmv <- cfa(model = cfa.model, 
               data = df2,
               estimator = "WLSMV",
               ordered = TRUE,
               rotation = "oblimin")

fit_metrics_robust <- c("chisq.scaled", "df", "pvalue.scaled", 
                         "cfi.robust", "tli.robust", "rmsea.robust", 
                        "rmsea.ci.lower.robust","rmsea.ci.upper.robust",
                        "srmr")

mfit1 <- 
  fitmeasures(fit.wlsmv, fit_metrics_robust) %>% 
  rbind() %>% 
  as.data.frame() %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>%
  rename(Chi2 = chisq.scaled,
         p = pvalue.scaled,
         CFI = cfi.robust,
         TLI = tli.robust,
         RMSEA = rmsea.robust,
         CI_low = rmsea.ci.lower.robust,
         CI_high = rmsea.ci.upper.robust,
         SRMR = srmr) %>% 
  add_column(Model = "WLSMV", .before = "Chi2")

kbl_rise(mfit1)

```

```{r}
modificationIndices(fit.wlsmv,
                    standardized = T) %>% 
  as.data.frame(row.names = NULL) %>% 
  #filter(mi > 30,
  #       op == "~~") %>% 
  arrange(desc(mi)) %>% 
  mutate(across(where(is.numeric),~ round(.x, 3))) %>%
  kbl_rise(fontsize = 14, tbl_width = 75)

```

## Software used

```{r}
pkgs <- cite_packages(cite.tidyverse = TRUE,
                      output = "table",
                      bib.file = "grateful-refs.bib",
                      include.RStudio = TRUE,
                      out.dir = getwd())
formattable(pkgs, 
            table.attr = 'class=\"table table-striped\" style="font-size: 15px; font-family: Lato; width: 80%"')

```

```{r}
sessionInfo()
```

## References
